---
title: "GST & Pricing in the Indian Motorcycle Market"
subtitle: "2020–2025"
author: "Vrunda Patil, Vaishnav Nayak, Sin Chan, Chenghuiyu Fan"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    footer: "ETC5513 Assignment 3 — Michael Rangers"
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup
library(tidyverse)
library(knitr)
library(kableExtra)
bikes <- read_csv("data/indian_bikes_dataset_1000.csv")
```

## Why does this matter?

Every time someone in India buys a motorcycle, they pay **28% GST on top of the price** — no exceptions.

But not everyone feels that equally.

This report asks: **who really bears the burden?**

## The Indian Motorcycle Market

India sells **millions of two-wheelers every year** — it is the largest market in the world.

Three very different types of buyers exist:

| Segment    | Who buys it                            |
|------------|----------------------------------------|
|  Budget    | Daily commuters, low-income households |
|  Mid-range | Middle-income urban buyers             |
|  Premium   | Enthusiasts, higher-income buyers      |

All three pay the **same 28% GST rate.**

## Our Dataset

We used a publicly available dataset from Kaggle covering **1,000 Indian motorcycles** from **2020 to 2025.**

- **8 brands** — Honda, Bajaj, Hero, Royal Enfield, TVS, Yamaha, KTM, Suzuki
- **18 variables** — prices, GST, engine size, mileage, buyer behavior and price sensitivity
- **Clean data** — ready to analyse in R

A five-year window that includes COVID-19 disruption, supply chain shortages and rising prices.

## How we analysed it

We kept it straightforward:

-   Grouped all 1,000 bikes by **budget, mid and premium** segment
-   Calculated average prices, GST amounts and engine specs
-   Looked at whether **price sensitivity changes the GST burden**
-   Built everything in **R** using `tidyverse` and `kableExtra`
-   The full report and these slides are both **reproducible from the same GitHub repository**

```{r}
#| label: tbl-methods
#| tbl-cap: "What each segment looks like on average"
bikes |>
  group_by(segment) |>
  summarise(
    `Models` = n(),
    `Avg Engine (cc)`  = round(mean(cc), 1),
    `Avg On-Road (INR)` = round(mean(on_road_price_inr), 0),
    `Avg GST (INR)` = round(mean(gst_amount_inr), 0)
  ) |>
  rename(Segment = segment) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(font_size = 22, full_width = FALSE)
```

## What the numbers show

The price gap between segments is enormous — and GST makes it bigger.

```{r}
#| label: fig-price
#| fig-width: 9
#| fig-height: 4.5
bikes %>%
  group_by(segment) %>%
  summarise(
    Factory_Price = mean(factory_price_inr),
    GST_Amount = mean(gst_amount_inr),
    On_Road_Price = mean(on_road_price_inr),
    Other_Charges = On_Road_Price - (Factory_Price + GST_Amount),
    .groups = "drop"
  ) %>%
  
  pivot_longer(
    cols = c(Factory_Price, GST_Amount, Other_Charges),
    names_to = "Component", values_to = "Amount"
  ) %>%
  
  mutate(Component = factor(Component,
    levels = c("Other_Charges", "GST_Amount", "Factory_Price"))) %>%
  
  ggplot(aes(x = segment, y = Amount, fill = Component)) +
  geom_bar(stat = "identity", position = "stack", width = 0.5) +
  scale_fill_brewer(palette = "Set2",
    labels = c("Other Charges", "GST Amount", "Factory Price")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Market Segment", y = "Price (INR)", fill = "Component") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

Budget buyers pay **INR 18,032** in GST. Premium buyers pay **INR 62,207** — more than three times as much.

## Does being price-sensitive help?

You might expect that buyers who are more price-sensitive pay less somehow. They do not.

```{r}
#| label: fig-sensitivity
#| fig-width: 9
#| fig-height: 4.2
bikes |>
  mutate(price_sensitivity = factor(price_sensitivity)) |>
  ggplot(aes(x = price_sensitivity,
             y = gst_amount_inr,
             fill = segment)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(
    values = c("budget"= "#2c7bb6",
               "mid"= "#f39c12",
               "premium" = "#c0392b"),
    labels = c("Budget", "Mid", "Premium")
  ) +
  labs(x = "Price Sensitivity Level",
       y = "GST Amount (INR)",
       fill = "Market Segment") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

The boxes are flat across all sensitivity levels. **Your GST is decided by your segment — nothing else.**

## What this means in practice

::: incremental
-   A budget buyer earning less pays **INR 18,032** in tax
-   A premium buyer earning more pays **INR 62,207** in tax
-   Both face the **exact same 28% rate**
-   The budget buyer feels it **much more** relative to their income
-   The flat rate creates **structurally unequal pressure** across the whole market
:::

## What should change?

**For manufacturers:** Budget brands should avoid passing costs to buyers — add value through features or service instead.

**For premium brands:** Keep watching price sensitivity data — it can shift quickly.

**For policymakers:** A single 28% rate sounds fair but is not in practice. Differentiated rates by engine capacity could reduce the burden on everyday commuters without disrupting the premium market.

## References {.smaller}

::: {#refs}
:::
